### 2013-09-04 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
fixing some more typos in the nonlinear chapter of the document

\subsection{Inexact Newton Methods}
\label{subsec:inexact_newton_methods}
Inexact Newton methods arise when the Jacobian operator is not exactly
inverted as initially described by Dembo and others
\cite{dembo_inexact_1982}. For common sparse nonlinear systems, which
in turn yield a sparse Jacobian matrix, this situation occurs when
conventional iterative methods are applied. In their definition, Dembo
formulated inexact methods such that they are independent of the
linear method used to solve for the Newton correction and therefore
are amenable to use with any linear solver. Furthermore, they bind the
convergence of the outer nonlinear iteration to the inner linear
iteration such that:
\begin{equation}
  ||\ve{J}(\ve{u}^k)\delta \ve{u}^k + \ve{F}(\ve{u}^k)|| \leq \eta^k
  ||\ve{F}(\ve{u}^k)||\:,
  \label{eq:inexact_newton_forcing}
\end{equation}
where $\eta^k \in [0,1)$ is defined as the \textit{forcing term} at
  nonlinear iteration $k$, the right-hand side is the nonlinear
  residual at nonlinear iteration $k$, and the left-hand side is the
  residual for the linear model at nonlinear iteration
  $k$. Eq~(\ref{eq:inexact_newton_forcing}) then states that the
  residual generated by the linear solver is bound by the nonlinear
  residual and how tightly it is bound is defined by the forcing
  term. This is useful in that we can vary how tightly coupled the
  convergence of the linear iterations used to generate the Newton
  correction is to the nonlinear iteration by relaxing or tightening
  the convergence properties on the linear iterative method. As a
  result, strategies for determining the forcing term can vary
  depending on the problem type and can greatly affect the convergence
  of the method or even prohibit convergence
  \cite{eisenstat_choosing_1996}. In addition, \textit{globalization
    methods} may be used to modify the Newton correction in a more
  desirable direction such that nonlinear convergence properties can
  be improved when the initial guess for $\ve{u}$ is poor
  \cite{pawlowski_globalization_2006}.

\subsection{Newton-Krylov Methods\ }
\label{subsec:newton_krylov_methods}
A form of inexact Newton methods, \textit{Newton-Krylov methods} are
nonlinear iterative methods that leverage a Krylov subspace method as
the linear solver for generating the Newton correction
\cite{kelley_iterative_1995}. As outlined in
Appendix~\ref{ch:linear_problem}, Krylov methods are robust and enjoy
efficient parallel implementations on modern
architectures. Furthermore, their lack of explicit dependence on the
operator make them easier to implement than other methods. In most
nonlinear problems, the Jacobian operator is generally non-symmetric
and therefore either Krylov methods with long recurrence relations
that can handle non-symmetric systems must be considered or the Newton
correction system must be preconditioned such that the operator is
symmetric and short recurrence relation methods can be potentially be
used.

@@

Upon preliminary inspection, FANM is simply a variant of an inexact
Newton method with a specific requirement of a fully formed
Jacobian. However, FANM has the potential to provide value in several
areas. First, if iterative or parallel performance of MCSA can be
demonstrated to better than that of GMRES or similar subspace methods
for asymmetric systems, FANM may outperform a Newton-Krylov
method. Second, for situations where memory is a constraint,
especially in situations where the Jacobian is generated often if not
at every Newton iteration, not building a subspace for ill-conditioned
problems will alleviate the associated cost. However, until the
preconditioning issues associated with MCSA as presented in the
previous chapter using the $SP_N$ equations can be resolved, this
hypothesis can unfortunately not be researched. Explicit algebraic
preconditioning used in conventional solution methods will almost
certainly be required to reduce the spectral radius of most Jacobian
operators for MCSA to apply and therefore the current preconditioning
strategy will eliminate all potential memory benefits of FANM.

@@

\subsubsection{Backward-Facing Step Problem}
\label{subsubsec:backward_facing_step}
The third benchmark was generated by Gartling in 1990 and consists of
both flow over a backward step and an outflow boundary condition
\cite{gartling_test_1990}. Using the mass and momentum equations
while neglecting the energy equation, this problem utilizes a longer
domain with a 1/25 aspect ratio used for this work\footnote{The
  literature on this benchmark suggests using a domain with a 1/30
  aspect ratio to reduce the effect of the outflow boundary condition
  on the upstream recirculation zones. Using a 1/25 aspect ratio for
  this work was not observed to disturb the recirculation zones and
  was therefore used to reduce memory pressure on the FANM solver.}
with the boundary conditions as shown in
Figure~\ref{fig:backward_facing_step}.

@@

A Drekar model of thermal convection cavity problem was solved using
both the Newton-Krylov solver and the FANM solver for Rayleigh numbers
of \sn{1}{3}, \sn{1}{4}, \sn{1}{5} and \sn{1}{6} on a $50 \times 50$
square grid. Table~\ref{tab:thermal_convection_parameters} gives the
physical parameters of the problem and
Table~\ref{tab:convection_mcsa_parameters} gives the MCSA solver
parameters used within the FANM method. Variation in the Rayleigh
number of the system was achieved by keeping the parameters given in
Table~\ref{tab:thermal_convection_parameters} constant and varying the
temperature of the hot side of the cavity. In addition to multigrid
preconditioning for the linear solution at each Newton step,
backtracking as outlined in Shadid's work \cite{shadid_inexact_1997}
was used as a means of globalization to assist the Newton correction
computation while the forcing term at all Rayleigh numbers but
\sn{1}{6} was computed at each Newton step using method 2 from
Shadid's work. In this forcing term method, nonlinear residuals from
the two previous Newton steps are used to inform the choice of the
forcing term for the next step. At a Rayleigh number of \sn{1}{6}, a
constant forcing term of \sn{1}{-4} was use to give better convergence
properties.

@@

A Drekar model of thermal convection cavity problem was solved using
both the Newton-Krylov solver and the FANM solver for Reynolds numbers
of 100, 300, 500, and 700 on a $50 \times 50$ square
grid. Table~\ref{tab:lid_driven_parameters} gives the physical
parameters of the problem and Table~\ref{tab:driven_mcsa_parameters}
gives the MCSA solver parameters used within the FANM
method. Variation in the Reynolds number of the system was achieved by
keeping the parameters given in Table~\ref{tab:lid_driven_parameters}
fixed and varying the fluid velocity in the $x$ direction on the top
boundary. In addition to multigrid preconditioning for the linear
solution at each Newton step, a polynomial line search method was used
as a means of globalization, assisting the Newton correction
computation as outlined by Pawlowski in
\cite{pawlowski_globalization_2006} while the forcing term was again
computed using method 2 from Shadid's work \cite{shadid_inexact_1997}
at all Reynolds numbers.

@@

A Drekar model of the backward facing step problem was solved using
both the Newton-Krylov solver and the FANM solver for Reynolds numbers
of 200, 300, 400, and 500 on a $20 \times 250$ rectangular
grid. Table~\ref{tab:backward_step_parameters} gives the physical
parameters of the problem and Table~\ref{tab:step_mcsa_parameters}
gives the MCSA solver parameters used within the FANM
method. Variation in the Reynolds number of the system was achieved by
keeping the parameters given in
Table~\ref{tab:backward_step_parameters} fixed and varying the maximum
fluid velocity in the parabolic inlet condition. In addition to
multigrid preconditioning for the linear solution at each Newton step,
a polynomial line search method was used as a means of globalization
while the forcing term was again computed using method 2 from Shadid's
work \cite{shadid_inexact_1997}.

@@

In addition to nonlinear iterative performance, performance of the
inner linear solves is also of interest. Here, we seek to minimize the
minimum number of total linear solver iterations over the course of
all nonlinear iterations in order to limit matrix-vector multiplies
and other costly
operations. Table~\ref{tab:convection_linear_iter_comparison} gives
the total number of GMRES, MCSA, and Richardson iterations required to
converge the Newton method at each Rayleigh number used. For all
cases, fewer MCSA iterations were required than GMRES iterations to
converge the problem even with both solvers preconditioned with the
same multigrid package and the same parameters. For the \sn{1}{6}
Rayleigh number case where a constant forcing term of \sn{1}{-4} was
kept, this difference is most drastic with FANM requiring only 64\% of
the linear iterations required by the Newton-Krylov scheme with the
linear solver per nonlinear iteration given for this case in
Figure~\ref{fig:ra1e6_linear_iters}. Even the \sn{1}{5} Rayleigh
number case that required more Newton iterations required 5 fewer MCSA
iterations to converge than GMRES iterations. In all cases, MCSA
improved on the iterative performance of using only the Richardson
iteration through the residual Monte Carlo acceleration. If the
performance of the MCSA implementation used here were to be
potentially optimized in the future, this improved iterative
performance could also mean improved time to solution.

@@

The backward step problem exhibited different performance
characteristics than the other two benchmark problems, likely due to
the different geometry and boundary
conditions. Table~\ref{tab:step_nonlinear_iter_comparison} gives the
nonlinear iteration results for this benchmark. At low Reynolds
numbers, FANM converges in fewer nonlinear iterations than the
Newton-Krylov solver. However, as the problem becomes more difficult
to solver at higher rates of flow, the Newton-Krylov solver eventually
converges in fewer nonlinear iterations. We see the same behavior as
well in the linear solver iteration data given by
Table~\ref{tab:step_linear_iter_comparison}. At the 3 lowest Reynolds
numbers, FANM required significantly fewer linear solver iterations to
converge the problem while the ill-conditioned case at a Reynolds
number of 500 is observed to use many more MCSA iterations. For
Reynolds numbers of 200, 300, and 500, FANM also reduced the number of
Newton steps need when compared to the Newton-Richardson scheme.

@@

For the total linear solver iterations given in
Table~\ref{tab:benchmark_linear_comparison}, the difference is more
dramatic with FANM converging the nonlinear residual in fewer linear
solver iterations for twice as many cases as the Newton-Krylov method
keeping in mind that for all cases both GMRES and MCSA were
preconditioned with identical multigrid parameters using the same
multigrid preconditioning library. This shows that when properly
preconditioned, we can expect performance improvements in terms of
total linear solver iterations required to converge the nonlinear
residual. With future optimization of MCSA implementations, this may
also mean improved time to solution for FANM when compared to
Newton-Krylov methods.

>>>>

\subsection{Inexact Newton Methods}
\label{subsec:inexact_newton_methods}
Inexact Newton methods arise when the Jacobian operator is not exactly
inverted as initially described by Dembo and others
\cite{dembo_inexact_1982}. For common sparse nonlinear systems, which
in turn yield a sparse Jacobian matrix, this situation occurs when
conventional iterative methods are applied. In their definition, Dembo
formulated inexact methods such that they are independent of the
linear method used to solve for the Newton correction and therefore
are amenable to use with any linear solver. Furthermore, they bind the
convergence of the outer nonlinear iteration to the inner linear
iteration such that:
\begin{equation}
  ||\ve{J}(\ve{u}^k)\delta \ve{u}^k + \ve{F}(\ve{u}^k)|| \leq \eta^k
  ||\ve{F}(\ve{u}^k)||\:,
  \label{eq:inexact_newton_forcing}
\end{equation}
where $\eta^k \in [0,1)$ is defined as the \textit{forcing term}, the
  right-hand side is the nonlinear residual, and the left-hand side is
  the residual for the linear model with all defined at nonlinear
  iteration $k$. Eq~(\ref{eq:inexact_newton_forcing}) then states that
  the residual generated by the linear solver is bound by the
  nonlinear residual and how tightly it is bound is defined by the
  forcing term. As a result, strategies for determining the forcing
  term can vary depending on the problem type and can greatly affect
  the convergence of the method or even prohibit convergence
  \cite{eisenstat_choosing_1996}. In this work we will either use a
  constant forcing term or use method 2 from Shadid's work
  \cite{shadid_inexact_1997}:
\begin{equation}
  \eta_k = \gamma \Big(
  \frac{||\ve{F}(\ve{u}_k)||}{||\ve{F}(\ve{u}_{k-1})||} \Big)^{\alpha}\:,
  \label{eq:shadid_forcing}
\end{equation}
where $\gamma$ and $\alpha$ are free parameters with typical values of
0.9 for $\gamma$ and 2.0 for $\alpha$. In addition,
\textit{globalization methods} may be used to modify the Newton
correction in a more desirable direction such that nonlinear
convergence properties can be improved when the initial guess for
$\ve{u}$ is poor \cite{pawlowski_globalization_2006}.

\subsection{Newton-Krylov Methods\ }
\label{subsec:newton_krylov_methods}
A form of inexact Newton methods, \textit{Newton-Krylov methods} are
nonlinear iterative methods that leverage a Krylov subspace method as
the linear solver for generating the Newton correction
\cite{kelley_iterative_1995}. As outlined in
Appendix~\ref{ch:linear_problem}, Krylov methods are robust and enjoy
efficient parallel implementations on modern
architectures. Furthermore, their lack of explicit dependence on the
operator make them easier to implement than other methods. In most
nonlinear problems, the Jacobian operator is generally non-symmetric
and therefore either Krylov methodsthat can handle non-symmetric
systems must be considered or the Newton correction system must be
preconditioned such that the operator is symmetric.

@@

Upon preliminary inspection, FANM is simply a variant of an inexact
Newton method with a specific requirement of a fully formed
Jacobian. However, FANM has the potential to provide value in several
areas. First, if iterative or parallel performance of MCSA can be
demonstrated to be better than that of GMRES or similar subspace
methods for asymmetric systems, FANM may outperform a Newton-Krylov
method. Second, for situations where memory is a constraint,
especially in situations where the Jacobian is generated often if not
at every Newton iteration, not building a subspace for ill-conditioned
problems will alleviate the associated cost. However, until the
preconditioning issues associated with MCSA as presented in the
previous chapter using the $SP_N$ equations can be resolved, this
hypothesis can unfortunately not be researched. Explicit algebraic
preconditioning used in conventional solution methods will almost
certainly be required to reduce the spectral radius of most Jacobian
operators for MCSA to apply and therefore the current preconditioning
strategy will eliminate all potential memory benefits of FANM.

@@

\subsection{Backward-Facing Step Problem}
\label{subsubsec:backward_facing_step}
The third benchmark was generated by Gartling in 1990 and consists of
both flow over a backward step and an outflow boundary condition
\cite{gartling_test_1990}. Using the mass and momentum equations
while neglecting the energy equation, this problem utilizes a longer
domain with a 1/25 aspect ratio used for this work\footnote{The
  literature on this benchmark suggests using a domain with a 1/30
  aspect ratio to reduce the effect of the outflow boundary condition
  on the upstream recirculation zones. Using a 1/25 aspect ratio for
  this work was not observed to disturb the recirculation zones and
  was therefore used to reduce memory pressure on the FANM solver.}
with the boundary conditions as shown in
Figure~\ref{fig:backward_facing_step}.

@@

A Drekar model of thermal convection cavity problem was solved using
both the Newton-Krylov solver and the FANM solver for Rayleigh numbers
of \sn{1}{3}, \sn{1}{4}, \sn{1}{5} and \sn{1}{6} on a $50 \times 50$
square grid. Table~\ref{tab:thermal_convection_parameters} gives the
physical parameters of the problem and
Table~\ref{tab:convection_mcsa_parameters} gives the MCSA solver
parameters used within the FANM method. Variation in the Rayleigh
number of the system was achieved by keeping the parameters given in
Table~\ref{tab:thermal_convection_parameters} constant and varying the
temperature of the hot side of the cavity. In addition to multigrid
preconditioning for the linear solution at each Newton step,
backtracking as outlined in Shadid's work \cite{shadid_inexact_1997}
was used as a means of globalization to assist the Newton correction
computation while the forcing term at all Rayleigh numbers but
\sn{1}{6} was computed at each Newton step using
Eq~\ref{eq:shadid_forcing}. At a Rayleigh number of \sn{1}{6}, a
constant forcing term of \sn{1}{-4} was use to give better convergence
properties.

@@

A Drekar model of thermal convection cavity problem was solved using
both the Newton-Krylov solver and the FANM solver for Reynolds numbers
of 100, 300, 500, and 700 on a $50 \times 50$ square
grid. Table~\ref{tab:lid_driven_parameters} gives the physical
parameters of the problem and Table~\ref{tab:driven_mcsa_parameters}
gives the MCSA solver parameters used within the FANM
method. Variation in the Reynolds number of the system was achieved by
keeping the parameters given in Table~\ref{tab:lid_driven_parameters}
fixed and varying the fluid velocity in the $x$ direction on the top
boundary. In addition to multigrid preconditioning for the linear
solution at each Newton step, a polynomial line search method was used
as a means of globalization, assisting the Newton correction
computation as outlined by Pawlowski in
\cite{pawlowski_globalization_2006} while the forcing term was again
computed using Eq~\ref{eq:shadid_forcing} at all Reynolds numbers.

@@

A Drekar model of the backward facing step problem was solved using
both the Newton-Krylov solver and the FANM solver for Reynolds numbers
of 200, 300, 400, and 500 on a $20 \times 250$ rectangular
grid. Table~\ref{tab:backward_step_parameters} gives the physical
parameters of the problem and Table~\ref{tab:step_mcsa_parameters}
gives the MCSA solver parameters used within the FANM
method. Variation in the Reynolds number of the system was achieved by
keeping the parameters given in
Table~\ref{tab:backward_step_parameters} fixed and varying the maximum
fluid velocity in the parabolic inlet condition. In addition to
multigrid preconditioning for the linear solution at each Newton step,
a polynomial line search method was used as a means of globalization
while the forcing term was again computed using Eq~\ref{eq:shadid_forcing}.

@@

In addition to nonlinear iterative performance, performance of the
inner linear solves is also of interest. Here, we seek to minimize the
number of total linear solver iterations over the course of all
nonlinear iterations in order to limit matrix-vector multiplies and
other costly
operations. Table~\ref{tab:convection_linear_iter_comparison} gives
the total number of GMRES, MCSA, and Richardson iterations required to
converge the Newton method at each Rayleigh number used. For all
cases, fewer MCSA iterations were required than GMRES iterations to
converge the problem even with both solvers preconditioned with the
same multigrid package and the same parameters. For the \sn{1}{6}
Rayleigh number case where a constant forcing term of \sn{1}{-4} was
kept, this difference is most drastic with FANM requiring only 64\% of
the linear iterations required by the Newton-Krylov scheme with the
linear solver iterations at every nonlinear iteration given for this
case in Figure~\ref{fig:ra1e6_linear_iters}. Even the \sn{1}{5}
Rayleigh number case that required more Newton iterations required 5
fewer MCSA iterations to converge than GMRES iterations. In all cases,
MCSA improved on the iterative performance of using only the
Richardson iteration through the residual Monte Carlo acceleration. If
the performance of the MCSA implementation used here were to be
potentially optimized in the future, this improved iterative
performance could also mean improved time to solution.

@@

The backward step problem exhibited different performance
characteristics than the other two benchmark problems, likely due to
the different geometry and boundary
conditions. Table~\ref{tab:step_nonlinear_iter_comparison} gives the
nonlinear iteration results for this benchmark. At low Reynolds
numbers, FANM converges in fewer nonlinear iterations than the
Newton-Krylov solver. However, as the problem becomes more difficult
to solver at higher rates of flow, the Newton-Krylov solver eventually
converges in fewer nonlinear iterations. We see the same behavior as
well in the linear solver iteration data given by
Table~\ref{tab:step_linear_iter_comparison}. At the 3 lowest Reynolds
numbers, FANM required significantly fewer linear solver iterations to
converge the problem while the ill-conditioned case at a Reynolds
number of 500 is observed to use many more MCSA iterations. For
Reynolds numbers of 200, 300, and 500, FANM also reduced the number of
Newton steps needed when compared to the Newton-Richardson scheme.

@@

For the total linear solver iterations given in
Table~\ref{tab:benchmark_linear_comparison}, the difference is more
dramatic with FANM converging the nonlinear residual in fewer linear
solver iterations for twice as many cases as the Newton-Krylov method,
keeping in mind that for all cases both GMRES and MCSA were
preconditioned with identical multigrid parameters using the same
multigrid preconditioning library. This shows that when properly
preconditioned, we can expect performance improvements in terms of
total linear solver iterations required to converge the nonlinear
residual. With future optimization of MCSA implementations, this may
also mean improved time to solution for FANM when compared to
Newton-Krylov methods.
